---
title: "Getting Started with athleteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with athleteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
library(athleteR)
```

## Introduction

**athleteR** is an R package designed for sports scientists, coaches, and
analysts who need reproducible tools to explore and communicate athletic
performance data. It provides:

- S3 classes (`athlete`, `team`) for structured representation of sports data  
- Functions to compute performance scores, rank athletes, and measure consistency  
- Base-R visualizations (line charts, bar charts, radar plots)  
- Two built-in datasets for immediate exploration

---

## 1. Creating Athlete Objects

The foundational unit in **athleteR** is the `athlete` S3 object. Use
`new_athlete()` to create one:

```{r create-athletes}
a1 <- new_athlete(
  name     = "Jordan Smith",
  sport    = "basketball",
  position = "guard",
  age      = 26,
  stats    = c(points = 22.4, assists = 7.1, rebounds = 4.3),
  seasons  = 4
)

a2 <- new_athlete(
  name     = "Alex Rivera",
  sport    = "basketball",
  position = "forward",
  age      = 29,
  stats    = c(points = 18.8, assists = 4.2, rebounds = 9.5),
  seasons  = 7
)

print(a1)
```

`summary()` provides a statistical overview of the athlete's stats:

```{r summary-athlete}
summary(a1)
```

---

## 2. Building a Team

Group multiple `athlete` objects into a `team`:

```{r create-team}
a3 <- new_athlete("Morgan Lee", "basketball", "center", 31,
                  c(points = 14.0, assists = 2.9, rebounds = 11.2), 9)

squad <- new_team(
  name     = "Thunder Hawks",
  sport    = "basketball",
  athletes = list(a1, a2, a3),
  season   = "2025-26"
)

print(squad)
```

### Team Summary

`team_summary()` aggregates statistics across all roster members:

```{r team-summary}
team_summary(squad)
```

---

## 3. Performance Scoring

`performance_score()` computes a weighted composite metric. By default all
statistics are weighted equally (summed). Custom weights let you emphasise
the metrics most relevant to your sport.

```{r perf-score}
# Equal weighting
performance_score(a1)

# Prioritise points over other metrics
performance_score(a1, weights = c(points = 3, assists = 1.5, rebounds = 1))
```

---

## 4. Comparing Athletes

`compare_athletes()` returns a tidy data frame aligning shared statistics
side by side, making it easy to identify strengths and gaps:

```{r compare}
compare_athletes(a1, a2, a3)
```

---

## 5. Working with the Built-in Datasets

**athleteR** ships with two ready-to-use datasets.

### `sample_athletes` — cross-sectional

```{r sample-athletes}
data(sample_athletes)
head(sample_athletes)
```

Find the top 5 scorers:

```{r top5}
top_performers(sample_athletes, metric = "points", n = 5)
```

Normalize statistics to a 0–100 scale for fair comparison:

```{r normalize}
norm_df <- normalize_stats(sample_athletes, columns = c("points", "assists"))
head(norm_df[, c("name", "sport", "points", "assists")])
```

### `sample_seasons` — longitudinal

```{r sample-seasons}
data(sample_seasons)
head(sample_seasons)
```

Compute season-over-season growth:

```{r growth}
growth_df <- season_growth(sample_seasons, "name", "season", "points")
head(growth_df[!is.na(growth_df$growth_pct), ])
```

Measure consistency (lower CV = more consistent):

```{r consistency}
consistency_index(sample_seasons, id_col = "name", stat_col = "points")
```

---

## 6. Visualizations

### Performance Over Time

```{r plot-perf, fig.cap="Season-by-season points for selected athletes"}
plot_performance(
  data       = sample_seasons,
  id_col     = "name",
  season_col = "season",
  stat_col   = "points",
  athletes   = unique(sample_seasons$name)[1:4],
  title      = "Points Per Season (2018–2025)"
)
```

### Top Performers Bar Chart

```{r plot-top, fig.cap="Top 8 athletes by points"}
plot_top_performers(sample_athletes, metric = "points", n = 8,
                    title = "Top 8 Scorers", color = "steelblue")
```

### Radar Chart Comparison

```{r radar, fig.cap="Head-to-head radar comparison"}
plot_radar(a1, a2, title = "Jordan Smith vs Alex Rivera")
```

---

## 7. Summary of Functions

| Function | Description |
|---|---|
| `new_athlete()` | Create an `athlete` S3 object |
| `new_team()` | Create a `team` S3 object |
| `performance_score()` | Weighted composite score |
| `compare_athletes()` | Side-by-side stat comparison |
| `team_summary()` | Aggregate team statistics |
| `top_performers()` | Rank athletes by a metric |
| `normalize_stats()` | Min-max normalize to 0–100 |
| `season_growth()` | Season-over-season growth % |
| `consistency_index()` | Coefficient of variation |
| `plot_performance()` | Line chart over seasons |
| `plot_top_performers()` | Horizontal bar chart |
| `plot_radar()` | Radar/spider comparison chart |

---

## Session Info

```{r session-info}
sessionInfo()
```
